---
output: github_document
---

# labellexicon

`labellexicon` is an **R** package for labelling factor variables in **R** data
frames. This is especially useful if you have encoded variables and want to 
plot your data with meaningful labels. Therefore, the `labellexicon` relabels
your variables and the label mappings are stored in **yaml files**, so called
**Lexicon** files. This makes
it very easy to use the same label mapping in multiple projects that have
similar data sets. 

Furthermore, `LabelLexicon` allows you to process your LabelLexicons in a 
way that is very similar to [dplyr]. You can `select`, `mutate` and `rename`
the entries of your LabelLexicon. You may also `merge` two or more LabelLexicas
into a single LabelLexicon. 

## Installation

```{r, eval = FALSE}
# Install development version from GitHub
devtools::install_github('a-maldet/labellexicon', build_opts = NULL)
```

## Usage

### Create a `LabelLexicon` from a list

You can simply create a LabelLexicon from a named list object, holding named character vectors.
Each entry of the named list represents a variable of your data and each named
character vector is a translation mapping for your variable. The names of the
character vectors represent the old values of the variable and the values of
the character vectors are the new labels that should be used.
```{r}
obj <- list(
  supp = c(VC = "Ascorbic acid", OJ = "Orange juice"),
  dose = c("0.5" = "Low", "1.0" = "Medium", "2.0" = "High")
)
```


### Load `LabelLexicon` file (yaml)

Structure of your Lexicon file `short_lex.yaml`
```{yaml}
supp:
  VC: Ascorbic acid
  OJ: Orange juice
dose:
  0.5: Low
  1.0: Medium
  2.0: High
```

```{r, include = FALSE}
library(labellexicon)
library(magrittr)
lex <- "data/short_lex.yaml" %>%
  read_lexicon_file
```

```{r, eval = FALSE}
library(labellexicon)
library(magrittr)
lex <- "short_lex.yaml" %>%
  read_lexicon_file
```

The label mappings are now contained in the LabelLexicon object `lex`.

```{r}
lex
```

### Relabel your variables with a LabelLexicon

Now, that we have loaded a LabelLexicon `lex`, we can use it in order 
to translate a categorical variable of our data.frame 
(not necessarily a factor variable) to a factor variable holding the labels
that are defined in the LabelLexicon `lex`.

```{r}
ToothGrowth %>%
  translate(lex, "supp") %>%
  translate(lex, "dose")
```
The variables `supp` and `dose` are now factor variables holding the new labels.

In case, your variable has a different name, than the entry in the LabelLexicon
you can use the `col` argument. If the resulting labled variable should be
stored under a differnt name, you may use the `col_new` argument.

```{r}
ToothGrowth %>%
  translate(lex, "supp") %>%
  translate(lex, "dose")
```

### Alter your LabelLexicon

Sometimes it can be useful to alter your LabelLexicon. With `select` you may
pick a subset of LabelLexicon entries.

```{r}
lex %>%
  select(c("supp", "dose"))
```

With `mutate` you can set a new label translation (character vector) for a variable:

```{r}
lex %>%
  mutate("supp", c(VC = "Ascorbic a.", OJ = "Orange j."))
```

With `rename` you can rename a LabelLexicon entry:

```{r}
lex %>%
  rename(c("supp", "dose"), c("SUPP", "DOSE"))
```

### Merge two ore more LabelLexicas

Sometimes you may want to merge two or more LabelLexicas into one LabelLexicon.
This can be done with ``

### Overriding error handlers in nested environments
The function `composerr_parent` looks up the existing error handler in the
parent environment and not in the current environment. This can be useful in
nested scoping situations (for example when checking if a nested list object
has a valid structure) and if you want to store the nested error handler 
functions under the same name (overriding error handler functions).

```{r, eval = FALSE}
### Example 2 ###
### check if all entries of the list object 'obj' are TRUE ###
obj <- list(x = TRUE, y = TRUE, z = FALSE)
# original error handler
err_h <- composerr("obj is invalid")
# check each list element 
sapply(names(obj), function(name) {
  # modify error handler to nested sitation
  err_h <- composerr_parent(paste("Error in", name), err_h)
  # check element and throw error FALSE
  if (!obj[[name]])
    err_h("Value is FALSE")
})
# --- resulting message ---
# "obj is invalid: Error in z: Value is FALSE"
```

## License

[GPL-3]
  
[GPL-3]: https://R-package.github.io/styledTables/LICENSE
[dplyr]: https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html
