---
output: github_document
---

# labelmachine

`labelmachine` is an **R** package that helps you assigning new labels to 
data.frame variables. Furthermore, you can manage your label translations
in so called **dictionary** files, which are **yaml** files.
This makes it very easy using the same label translations in multiple
projects that share similar data structure.

The most important functions are:

- `read_dictionary`: Reads in a **yaml** file holding the label translations
  for one or more variables. The function returns a **dictionary object**, that can be
  used for the translation of variable labels later on. 
- `translate`: Use a dictionary object in order to relabel one or more
  variables of your data.frame. The variables may be of type `number`,
  `character` or `factor`. In case of a factor variable, you have the possibility
  to keep the original ordering by using the 
  function argument `keep_ordering = TRUE`.
- `new_dictionary`: Create a dictionary object from a named list, holding the 
  translations for one or more variables.
- `write_dictionary`: Write a dictionary object to a **yaml** file.
- `select`: Pick a subset of translations in a dictionary object.
- `mutate`: Alter the translations in a dictionary
  object.
- `rename`: Rename a variable translations in a dictionary object.
- `merge`: Merge two or more dictionary objects into a single dictionary 
  object.

## Installation

```{r, eval = FALSE}
# Install development version from GitHub
devtools::install_github('a-maldet/labelmachine', build_opts = NULL)
```

## Usage

### Load a dictionary from a yaml file

Structure of your dictionary file `my_dictionary.yaml`:
```{yaml}
supp:
  VC: Ascorbic acid
  OJ: Orange juice
dose:
  "0.5": Low
  "1": Medium
  "2": High
```
The first level names are the names of the variables that should be translated.
The second level names represent
the original values of the variables and the third level holds the labels
that should be assigned to the original values of the variables.

Load your `dictionary` file with `read_dictionary`:
```{r, include = FALSE}
library(labelmachine)
library(magrittr)
dict <- "data/my_dictionary.yaml" %>%
  read_dictionary
```
```{r, eval = FALSE}
library(labelmachine)
library(magrittr)
dict <- "my_dictionary.yaml" %>%
  read_dictionary
```

### Assign new labels to your variables

You can use the dictionary object `dict` in order 
to translate categorical variables
(not necessarily factor variables) in your data.frame
to a factor variable holding the labels
that are defined in the dictionary `dict`.

Relable your data.frame variables with `translate`:
```{r, eval = FALSE}
# data.frame with original values 
ToothGrowth %>% head

# data.frame with new labels
ToothGrowth %>%
  translate(dict, c("supp", "dose")) %>%
  head
```

Now, the columns `supp` and `dose` are factor variables, which hold the
desired labels and have the same ordering as in the dictionary `dict`.
If the original variable is a factor variable and you want to keep
the original ordering, you can use the function argurment
`r keep_ordering = TRUE`.

### Create a dictionary object manually

Instead of reading in a yaml file you can also create a dictionary object manually
from a named list object, holding named character vectors.
Each entry of the named list represents a variable of your data.frame and
each named character vector is a translation. 
The names of the character vector entries represent the original values of
the variable and
the values of the character vector entries are the new labels that should be
assigned.

```{r}
dict <- list(
    supp = c(VC = "Ascorbic acid", OJ = "Orange juice"),
    dose = c("0.5" = "Low", "1.0" = "Medium", "2.0" = "High")
  ) %>%
  new_dictionary
```

### Alter your dictionary

Sometimes it can be useful to alter your dictionary object.

With `select` you may
pick a subset of dictionary entries:
```{r}
dict %>%
  select(c("supp", "dose"))
```

With `mutate` you can set a new translation (character vector)
for a variable:
```{r}
dict %>%
  mutate("supp", c(VC = "Ascorbic a.", OJ = "Orange j."))
```

With `rename` you can rename a dictionary entry:
```{r}
dict %>%
  rename(c("supp", "dose"), c("SUPP", "DOSE"))
```

### Merge two ore more LabelLexicas

With `merge` you can merge two or more LabelLexicas into one dictionary
object:

```{r}
dict1 <- list(
    gender = c("0" = "female", "1" = "male"),
    age = c("0" = "<10y", "1" = ">=10y"),
    country = c("0" = "Austria", "1" = "Australia")
  ) %>%
  new_dictionary
dict2 <- list(
    school = c("0" = "Primary", "1" = "Secondary"),
    gender = c("0" = "Girl", "1" = "Boy")
  ) %>%
  new_dictionary
merge(dict1, dict2)
```

The `merge` merges the arguments from left to right. Therefore, the 
variable translations of `dict1` are overwritten by the entries of `dict2`.
In this example the resulting dictionary has contains the variable
translations for `age` and `country` as defined in `dict1`
and the variable translations for `school` and `gender` as defined in `dict2`.
The variable translations for `gender` in `dict1` are overwritten by `dict2`.

## License

[GPL-3]
  
[GPL-3]: https://R-package.github.io/styledTables/LICENSE
[dplyr]: https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html
