---
output: github_document
---

# labellexicon

`labellexicon` is an **R** package that helps you assigning new labels to 
data.frame variables. Furthermore, you can manage your label translations
in so called **LabelLexicon** files, which are **yaml** files, which hold
the mappings (translations) of your variable labels.
This makes it very easy using the same label translations in multiple
projects that share similar data structure.

The most important functions are:

- `read_lexicon_file`: Reads in a **yaml** file holding the label translations
  for one or more variables.
- `translate`: Use a LabelLexicon object in order to relabel one or more
  variables of your data.frame. The variables may be of type `number`,
  `character` or `factor`. In case of a factor variable, you have the possibility
  to keep the original ordering by using the 
  function argument `keep_ordering = TRUE`.
- `new_lexicon`: Create a LabelLexicon object from a named list, holding the 
  label translations for one or more variables.
- `write_lexicon_file`: Write a LabelLexicon object to a **yaml** file.
- `select`: Pick a subset of a LabelLexicon object.
- `mutate`: Alter the translations of a variable in a LabelLexicon
  object.
- `rename`: Rename a variable in a LabelLexicon object.
- `merge`: Merge two or more LabelLexicon objects into a single LabelLexicon 
  object.

## Installation

```{r, eval = FALSE}
# Install development version from GitHub
devtools::install_github('a-maldet/labellexicon', build_opts = NULL)
```

## Usage

### Load LabelLexicon file (yaml)

Structure of your LabelLexicon file `short_lex.yaml`:
```{yaml}
supp:
  VC: Ascorbic acid
  OJ: Orange juice
dose:
  "0.5": Low
  "1": Medium
  "2": High
```
The first level names are the variable names. The second level names represent
the original values of the variables and the third level values are the labels
that should be assigned to the values of the variables.

Load your `LabelLexicon` file with `read_lexicon_file`:
```{r, include = FALSE}
library(labellexicon)
library(magrittr)
lex <- "data/short_lex.yaml" %>%
  read_lexicon_file
```

### Relabel your variables with a LabelLexicon

We can use the LabelLexicon object `lex` in order 
to translate a categorical variable 
(not necessarily a factor variable) in our data.frame
to a factor variable with the labels
that are defined in the LabelLexicon `lex`.

Relable your data.frame variables with `translate`:
```{r, eval = FALSE}
# data.frame with original values 
ToothGrowth %>% head

# data.frame with new labels
ToothGrowth %>%
  translate(lex, "supp") %>%
  translate(lex, "dose") %>%
  head
```

Now, the columns `supp` and `dose` are factor variables, which hold the
desired labels and have the same ordering as in the `LabelLexicon` file.
If the original variable is a factor variable and you want to keep
the original ordering, you can set the function argurment
`r keep_ordering = TRUE`.

### Create a LabelLexicon object manually

Instead of reading in a yaml file you can also create a LabelLexicon manually
from a named list object, holding named character vectors.
Each entry of the named list represents a variable of your data.frames and
each named character vector is a translation. 
The names of the character vector entries represent the original values of
the variable and
the values of the character vector entries are the new labels that should be
assigned.

```{r}
lex <- list(
    supp = c(VC = "Ascorbic acid", OJ = "Orange juice"),
    dose = c("0.5" = "Low", "1.0" = "Medium", "2.0" = "High")
  ) %>%
  new_lexicon
```

### Alter your LabelLexicon

Sometimes it can be useful to alter your LabelLexicon object.

With `select` you may
pick a subset of LabelLexicon entries:
```{r}
lex %>%
  select(c("supp", "dose"))
```

With `mutate` you can set a new label translation (character vector)
for a variable:
```{r}
lex %>%
  mutate("supp", c(VC = "Ascorbic a.", OJ = "Orange j."))
```

With `rename` you can rename a LabelLexicon entry:
```{r}
lex %>%
  rename(c("supp", "dose"), c("SUPP", "DOSE"))
```

### Merge two ore more LabelLexicas

Sometimes you may want to merge two or more LabelLexicas into one LabelLexicon.
This can be done with `merge_lexica`

### Overriding error handlers in nested environments
The function `composerr_parent` looks up the existing error handler in the
parent environment and not in the current environment. This can be useful in
nested scoping situations (for example when checking if a nested list object
has a valid structure) and if you want to store the nested error handler 
functions under the same name (overriding error handler functions).

```{r, eval = FALSE}
### Example 2 ###
### check if all entries of the list object 'obj' are TRUE ###
obj <- list(x = TRUE, y = TRUE, z = FALSE)
# original error handler
err_h <- composerr("obj is invalid")
# check each list element 
sapply(names(obj), function(name) {
  # modify error handler to nested sitation
  err_h <- composerr_parent(paste("Error in", name), err_h)
  # check element and throw error FALSE
  if (!obj[[name]])
    err_h("Value is FALSE")
})
# --- resulting message ---
# "obj is invalid: Error in z: Value is FALSE"
```

## License

[GPL-3]
  
[GPL-3]: https://R-package.github.io/styledTables/LICENSE
[dplyr]: https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html
