---
title: "Altering lama-dictionaries"
date: "`r Sys.Date()`"
author: "Adrian Maldet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Altering lama-dictionaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

The creation of lama-dictionaries was described in [Creating lama-dictionaries]
and in [Translating variables] we saw how to use them in order to assign
the right labels to our data.

Now we will see, how to effectively alter lama-dictionaries, so that
we get dictionaries holding the right translations. `labelmachine` contains
a light weight frame work for altering lama-dictionaries, similar to the 
package [dplyr]:

* `lama_rename()` and `lama_rename_()`: Renaming translations
* `lama_select()` and `lama_select_()`: Selecting a subset of translations
* `lama_mutate()` and `lama_mutate_()`: Altering one or more translations
* `lama_merge()`: Merging two or more dictionaries into one

The commands which have no lodash at the end of the command name
(`lama_rename()`, `lama_select()`, `lama_mutate()`) use
**non-standard evaluation**.
This means, that instead of passing in translation names as
strings (e.g. `lama_rename_(dict, "old", "new")`), we can pass in
**unquoted expressions** (e.g. `lama_rename(dict, new = old)`), which are
automatically parsed. Often non-standard evaluation saves some time in writing,
but sometimes we want to pass in the names as character vectors. 
In this case, we need to use the **standard evaluation** variants of the commands.
These commands have the same names, but end on a lodash 
(e.g. `lama_rename_()`, `lama_select_()`, `lama_mutate_()`).

In the following part we will alter the following dictionary:

```{r}
library(labelmachine)
dict <- new_lama_dictionary(
  sub = c(eng = "English", mat = "Mathematics", gym = "Gymnastics"),
  lev = c(b = "Basic", a = "Advanced"),
  result = c(
    "1" = "Good",
    "2" = "Passed",
    "3" = "Not passed",
    "4" = "Not passed",
    NA_ = "Missed",
    "0" = NA
  )
)
dict
```

## Rename translations

With the commands `lama_rename()` and `lama_rename_()` we can rename one or
more translations in a lama-dictionary.
With `lama_rename()` we can use unquoted expressions:
```{r}
dict_new <- lama_rename(
  dict,
  subject_new = sub,
  level_new = lev,
  result_new = result
)
dict_new
```

With `lama_rename_()` we can pass two character vectors. One character vector
holding the old translation names and one vector which contains the
new translation names which should be used:
```{r}
dict_new <- lama_rename_(
  dict,
  old = c("sub", "lev", "result"),
  new = c("subject_new", "level_new", "result_new")
)
dict_new
```

## Select a subset of translations

Sometimes we want to keep a subset of translations, in this case we can use
`lama_select()` and `lama_select_()`.
With `lama_select()` we can use unquoted translation names:
```{r}
dict_new <- lama_select(dict, sub, lev)
dict_new
```
The resulting dictionary `dict_new` now only contains the translations `sub`
and `lev`.

With `lama_select_()` we pass in a single character vector, which
holds the names of the translations we want to keep:
```{r}
dict_new <- lama_select_(dict, c("sub", "lev"))
dict_new
```

## Alter translations

The commands `lama_mutate()` and `lama_mutate_()` are used to alter or delete
existing translations in a lama-dictionary or add
new translations (named character vectors) to it.
With `lama_mutate()` we use unquoted expressions:
```{r}
dict_new <- lama_mutate(
  .data = dict,
  teacher = c(jane = "Jane Doe", john = "John Doe"),
  sub = c(geo = "Geography", sub),
  lev = NULL,
  result = c(P = "Passed", F = "Failed")
)
dict_new
```

Besides the argument `.data` all other arguments are translation assignment and
the given argument names are used as the names to which the
translations, given on the right hand side of the equation, will be assigned:

* The translation `c(jane = "Jane Doe", john = "John Doe")` is assigned to the
  translation name `teacher`.
* In the expression given for translation `sub`, uses the object name `sub`
  inside of the expression (e.g. `c(geo = "Geography", sub)`) and is evaluated
  in this way. Therefore, the resulting translation `sub` is the combination
  of the label assignment `geo = "Geographry"` and the label assignments
  given in the old translation `sub` (e.g. `c(eng = "English",
  mat = "Mathematics", gym = "Gymnastics")`).
* The assignment `lev = NULL` deletes the translation with the name `lev`.
* The assignment `result = c(P = "Passed", F = "Failed")` overwrites the 
  translation `result` with a new translation.

The command `lama_mutate_()` is uses standard evaluation and can only
alter one translation at a time. We pass in a character string holding the
name of the translation we want to alter and a second argument holding the
translation (named character vector), we want to assign:

```{r}
dict_new <- lama_mutate_(
  .data = dict,
  key = "result",
  translation = c(P = "Passed", F = "Failed")
)
dict_new
```

## Merging lama-dictionaries

With the command `lama_merge` we can merge two or more
lama-dictionaries together into a single lama-dictionary.

Let us consider the following dictionaries:
```{r}
dict_a <- new_lama_dictionary(a = c(a = "A"), x = c(x = "A"), y = c(y = "A"))
dict_b <- new_lama_dictionary(b = c(b = "B"), x = c(x = "B"), z = c(z = "B"))
dict_c <- new_lama_dictionary(c = c(c = "C"), z = c(x = "B"))
```

We merge them together into a new dictionary:
```{r}
dict_new <- lama_merge(dict_a, dict_b, dict_c, show_warnings = FALSE)
dict_new
```

The merging is done from left to right. This means that the lama-dictionary
`dict_a` is partially overwritten by `dict_b` and the resulting lama-dictionary
is then partially overwritten by `dict_c`.

## Further reading

* [Creating lama-dictionaries]
* [Translating variables]
* [Get started]

[Get started]: https://a-maldet.github.io/labelmachine/articles/labelmachine.html
[Creating lama-dictionaries]: https://a-maldet.github.io/labelmachine/articles/create_dictionaries.html
[Altering lama-dictionaries]: https://a-maldet.github.io/labelmachine/articles/alter_dictionaries.html
[Translating variables]: https://a-maldet.github.io/labelmachine/articles/translate.html
[dplyr]: https://dplyr.tidyverse.org

